\bfseries \boldmath \scititle

### 2025-12-20 latex/main.tex
Update abstract

\begin{document}

\maketitle

%%% The summary (abstract) should consist of a single paragraph of 150 words or
%%% fewer.

\section*{SUMMARY}

% NOTE: currently 150 words

Millions of SARS-CoV-2 genome sequences were collected 
during the COVID-19 pandemic, forming a dataset of unprecedented richness.
Estimated genealogies are fundamental to
understanding this ocean of data 
and form the primary input to many downstream analyses.
A basic assumption of methods to infer genealogies from viral 
genetic data is that recombination is negligible and 
the genealogy is a tree.
However, recombinant lineages have risen to global prevalence, 
and simple tree representations are therefore 
incomplete and potentially misleading.
We present sc2ts, a method to infer reticulate genealogies 
as an Ancestral Recombination Graph (ARG) in real time at pandemic scale.
We infer an ARG for 2.48 million SARS-CoV-2 genomes, 
which leverages the widely used tskit software ecosystem to support further analyses and 
visualisation.
This rich and validated resource clarifies
the relationships among recombinant lineages,
quantifies the rate of recombination over time,
and provides a 
lower bound on detectable recombination.

\section*{KEYWORDS}

%%%  Include up to 10 keywords, separated by commas.
%%%  Keywords entered in EM are not carried over; only
%%%  keywords included in the main text will be used
%%%  in the final article metadata. Please note that
%%%  for some journals, keywords are chosen by the editors.

Ancestral recombination graphs, recombination, SARS-CoV-2, ARG, tskit

\section*{INTRODUCTION}
% SHZ: The first paragraph should explain
% why recombination "can" be important in pathogens.
% Then, we explain why we need recombination-aware methods for pathogens.
% This should set the stage for why this work is really important,
% not just for SC2 but more broadly for the field of pathogen genomics.

% Classical phylo methods fundamentally do not apply to SARS-CoV-2 data, both
% in terms of the basic goal, assumptions about samples and how recombination
% can be treated.
Classical phylogenetics is largely concerned with inferring the evolutionary history 
of divergent species~\cite{Felsenstein2004}. Recombination among lineages, however, distorts the inference of species trees~\cite{Posada2002effect},
 biases downstream
analyses~\cite{Schierup2000consequences,Hedge2014bacterial}, and so requires careful consideration~\cite{Rannala2025recombination}.
Modern pathogen datasets are not sparsely sampled from different species
% Need to concrete. "large" means 100 to some people.
% SO: The phrase tens to hundreds of thousands seems odd though given the millions available for SARS-CoV-2: maybe "thousands to millions" 
% JK: How's that?
and may contain millions of genomes sampled from 
% JK: deleted point about recomb here as TB, eg, doesn't
a single species~\cite{
Shu2017,% GISAID ~270k FLU + others
Jolley2018open,% PubMLST, 34K genomes for e.g. Staph Auerus
Apetrei2021hiv,% HIV 1,026,217 sequences
Cryptic2022data,% TB 12k
Dyer2025enterobase% 1 million, various things like E-coli
}.
In contrast to the sophisticated model-based approaches
used to infer species trees~\cite{Suchard2018bayesian,Bouckaert2019beast,Minh2020iq},
comprehensive phylogenies of pathogen datasets are typically built with
simpler distance methods % citation for this?
or approximations~\cite{Price2009fasttree},
primarily due to computational scaling limitations.
While several methods have been proposed to incorporate
recombination explicitly into model-based phylogenetic
inference~\cite{Didelot2010inference,%
Vaughan2017inferring,%
Muller2020bayesian,%
Guo2022recombination,
Muller2022bayesian},
scalability
is a major issue and practical application has been limited.
Software support that explicitly \emph{uses}
recombination-aware inference is mostly restricted to
visualisation, for example of small-scale phylogenetic
networks~\cite{Huson2012,Vaughan2017,Solis-Lemus2017}
or tree changes along the genome~\cite{Rasmussen2023}.

>>>>

\begin{document}
\maketitle

% Abstract, in bold
% There are strict length limits, and not all formats have abstracts.
% Consult the journal instructions to authors for details.
% Do not cite any references in the abstract.
\begin{abstract}\bfseries
Millions of SARS-CoV-2 genome sequences were collected 
during the COVID-19 pandemic, forming a dataset of unprecedented richness.
Estimated genealogies are fundamental to
understanding this ocean of data 
and form the primary input to many downstream analyses.
A basic assumption of methods to infer genealogies from viral 
genetic data is that recombination is negligible and 
the genealogy is a tree.
However, recombinant lineages have risen to global prevalence, 
and simple tree representations are therefore 
incomplete and potentially misleading.
We present sc2ts, a method to infer reticulate genealogies 
as an Ancestral Recombination Graph (ARG) in real time at pandemic scale.
We infer an ARG for 2.48 million SARS-CoV-2 genomes, 
which leverages the widely used tskit software ecosystem to support further analyses and 
visualisation.
This rich and validated resource clarifies
the relationships among recombinant lineages,
quantifies the rate of recombination over time,
and provides a 
lower bound on detectable recombination.
\end{abstract}

% \section*{KEYWORDS}

% Ancestral recombination graphs, recombination, SARS-CoV-2, ARG, tskit

% SHZ: The first paragraph should explain
% why recombination "can" be important in pathogens.
% Then, we explain why we need recombination-aware methods for pathogens.
% This should set the stage for why this work is really important,
% not just for SC2 but more broadly for the field of pathogen genomics.

% Classical phylo methods fundamentally do not apply to SARS-CoV-2 data, both
% in terms of the basic goal, assumptions about samples and how recombination
% can be treated.

% The first paragraph of any Science paper does NOT have a heading
% Nor is it indented
\noindent
Classical phylogenetics is largely concerned with inferring the evolutionary history 
of divergent species~\cite{Felsenstein2004}. Recombination among lineages, however, distorts the inference of species trees~\cite{Posada2002effect},
 biases downstream
analyses~\cite{Schierup2000consequences,Hedge2014bacterial}, and so requires careful consideration~\cite{Rannala2025recombination}.
Modern pathogen datasets are not sparsely sampled from different species
% Need to concrete. "large" means 100 to some people.
% SO: The phrase tens to hundreds of thousands seems odd though given the millions available for SARS-CoV-2: maybe "thousands to millions" 
% JK: How's that?
and may contain millions of genomes sampled from 
% JK: deleted point about recomb here as TB, eg, doesn't
a single species~\cite{
Shu2017,% GISAID ~270k FLU + others
Jolley2018open,% PubMLST, 34K genomes for e.g. Staph Auerus
Apetrei2021hiv,% HIV 1,026,217 sequences
Cryptic2022data,% TB 12k
Dyer2025enterobase% 1 million, various things like E-coli
}.
In contrast to the sophisticated model-based approaches
used to infer species trees~\cite{Suchard2018bayesian,Bouckaert2019beast,Minh2020iq},
comprehensive phylogenies of pathogen datasets are typically built with
simpler distance methods % citation for this?
or approximations~\cite{Price2009fasttree},
primarily due to computational scaling limitations.
While several methods have been proposed to incorporate
recombination explicitly into model-based phylogenetic
inference~\cite{Didelot2010inference,%
Vaughan2017inferring,%
Muller2020bayesian,%
Guo2022recombination,
Muller2022bayesian},
scalability
is a major issue and practical application has been limited.
Software support that explicitly \emph{uses}
recombination-aware inference is mostly restricted to
visualisation, for example of small-scale phylogenetic
networks~\cite{Huson2012,Vaughan2017,Solis-Lemus2017}
or tree changes along the genome~\cite{Rasmussen2023}.
